---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data
### 1. Load the data
```{r}
#assuming that data has been downloaded and placed in working directory

if(!file.exists("activity.csv")) {
        unzip("activity.zip")
}
activity <- read.csv("activity.csv")
        
```

## What is mean total number of steps taken per day?
### 1. Calculate the total number of steps per day
```{r}
#use tapply to find the sum of steps for each day in the dataset

stepsperday <- with(activity, tapply(steps,date,sum,na.rm = TRUE))

#convert to data frame for use with ggplot
df <- data.frame(stepsperday)
```
### 2. Make a histogram of the total number of steps taken each day
```{r}
library(ggplot2)
ggplot(df, aes(x = stepsperday)) + 
        geom_histogram(binwidth = 5000, colour = "magenta", fill = "green") + 
        ggtitle("Histogram of total number of steps per day") + 
        xlab("Number of steps per day") + ylab("Count")
```
### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
#calculate mean and median number of steps per day for report
meanSteps <- mean(stepsperday)
medianSteps <- median(stepsperday)
```
The mean total number of steps per day is `r as.character(round(meanSteps))`.
The median total number of steps per day is `r as.character(medianSteps)`.

## What is the average daily activity pattern?
### 1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across days
```{r}
#calculate the average daily activity pattern with tapply and plot results
activityPattern <- with(activity, tapply(steps,interval,mean, na.rm = TRUE))
plot(x = names(activityPattern), y = activityPattern, type = "l", 
     xlab = "Interval", ylab = "Average number of steps", 
     main="Average number of steps across all days for each 5-minute interval")
```
### 2. Which 5-minute interval, on average across all days in the dataset, contains the maximum number of steps?
```{r}

#find the interval which on average contains the maximum number of steps
maxInterval <- max(activityPattern)
findNameOfmaxInterval <- activityPattern == maxInterval
maxIntervalName <- names(activityPattern[findNameOfmaxInterval])
```
The 5-minute interval that on average across all days contains the maximum number of steps is `r as.character(maxIntervalName)`.

## Imputing missing values
### 1. Calculate and report the total number of missing values in the dataset

```{r}
#calculate the total number of missing values in the dataset for reporting
activityNA <- is.na(activity$steps)
totalNA <- sum(activityNA)
```
The total number of missing values in the datasheet is `r as.character(totalNA)`.

### 2 & 3. Devise a strategy for filling in all of the missing values in the dataset. Create new dataset that is equal to original but with missing data filled in
```{r}
#impute value based on the mean for the 5-minute interval across all days
#using the previously calculated variable activityPattern. only impute value if
#value is NA. (use round for the mean since number of steps has to be a whole
#number). NewActivity is the variable where missing values are filled in.
NewActivity <- activity
for (i in 1:17568) {
        if (is.na(NewActivity$steps[i])) {
                interval <- NewActivity[i,3]
                find_mean <- names(activityPattern) == interval
                mean_interval <- activityPattern[find_mean]
                NewActivity$steps[i] <- round(mean_interval)
        }
}
NewActivity$steps <- as.integer(NewActivity$steps)
```
### 4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median
```{r}
#calculate total number of steps taken each day when missing values are imputed
#and place them in a data frame to be used in ggplot
stepsperday2 <- with(NewActivity, tapply(steps,date,sum, na.rm = FALSE))
df2 <- data.frame(stepsperday2)

ggplot(df2, aes(x = stepsperday2)) + 
        geom_histogram(binwidth = 5000, colour = "magenta", fill = "green") + 
        ggtitle("Histogram of total number of steps per day, with imputed values") + 
        xlab("Number of steps per day") + ylab("Count")

#calculate and report the mean and median total number of steps per day when
#missing values are imputed
meanSteps2 <- mean(stepsperday2)
medianSteps2 <- median(stepsperday2)

```
The mean total number of steps taken per day after imputing missing values is `r as.character(round(meanSteps2))`.
The median total number of steps taken per day after imputing missing values is `r as.character(medianSteps2)`.

## Are there differences in activity patterns between weekdays and weekends?
### 1 & 2. Create a new factor variable in the dataset with two levels (weekday and weekend) indicating whether a given date is a weekday or weekend day. Make a panel plot of the 5-minute interval and the average number of steps taken, averaged across all weekdays or weekends.
```{r}
#format the date variable and make new variable weekday using the weekdays
#function
NewActivity$date <- as.Date(NewActivity$date, format = "%Y-%m-%d")
NewActivity$weekday <- weekdays(NewActivity$date, abbreviate = FALSE)

#check if the day of week is a weekday or weekend and replace the value with
#either "weekday" or "weekend"
for (i in 1:17568) {
        if (NewActivity$weekday[i] == "Saturday" || 
            NewActivity$weekday[i] == "Sunday") {
                NewActivity$weekday[i] <- "weekend"
        }
        else { NewActivity$weekday[i] <- "weekday"
        }
}
#make the variable weekday into factor, aggregate number of steps by interval
#and weekday and calculate the mean. plot using ggplot and facet_wrap to
#seperate based on weekday/weekend
NewActivity$weekday <- as.factor(NewActivity$weekday)
byWeekdayInterval <- aggregate(steps ~ interval + weekday, NewActivity, mean)
ggplot(byWeekdayInterval, aes(interval,steps)) + geom_line(colour = "blue") + 
        facet_wrap(~weekday, ncol = 1) + ylab("number of steps")
```
